---
title: "Reversible jump for variable selection"
subtitle: "NIMBLE ENAR 2020 short course"
author: "NIMBLE Development Team"
output:
  html_document:
    code_folding: show
---


```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(methods)  # otherwise new() not being found
library(nimble)
```

# Introduction to Bayesian variable selection

- You have many candidate explanatory variables.
- Bayesian approach is to have a probability that a variable is included in the model.
- Really this is a probability that the coefficient is $\ne 0$.
- BUGS/JAGS implementation is with indicator variables.

```
  linear_predictor[i] <- beta0 + ind * beta1 * x[i]
```

- This has problems: when `ind` is 0, `beta1` follows its prior, until it hits a reasonable value for `beta1` that allows `ind` equal to 1 to be accepted.
- "Solution": informative priors

# Solution!: Reversible Jump MCMC

 - RJMCMC is a method for sampling across different models.
 - Specifically it is about sampling between different numbers of dimensions.
 - In full generality, RJ requires one to figure out a way to propose reasonable parameter values when moving between models. Hard!
 - RJ for variable selection is relatively simple.

    - We don't change the actual NIMBLE model object, but we turn on and off which dimensions are sampled.
    - Implementation, like all samplers, is written using `nimbleFunction`s.

Recall that we had two covariates in the AFT model - blood urea nitrogen and hemoglobin. Let's see if hemoglobin is needed in the model using Bayesian variable selection.

# RJMCMC for variable selection in nimble


- Update an MCMC configuration to use RJMCMC.



```{r, aft-rjmcmc-setup}

model_sel <- model$newModel(replicate = TRUE)  # or we could just `nimbleModel` again with the original model code
cmodel_sel <- compileNimble(model_sel)

conf_sel <- configureMCMC(model_sel)
configureRJ(conf_sel,
            targetNodes = 'delta[2]',
            priorProb = 0.5,
            control = list(mean = 0, scale = 2))
mcmc_sel <- buildMCMC(conf_sel)

cmcmc_sel <- compileNimble(mcmc_sel, project = model_sel)

# Running the MCMC

Now let's run the MCMC. We'll need to manually assign our special samplers for the RJ aspect of things. 
      
```{r, aft-mcmc}
resultsAFT <- runMCMC(cmcmc_sel, niter = 21000, nburnin = 100) ## burnin is post-thinning (to be fixed in next NIMBLE release)
```

Let's look at the MCMC behavior of the coefficient of interest.

```{r, results, fig.width=8, fig.height=5, fig.cap=''}
par(mfrow = c(1,2))
ts.plot(resultsAFT[ , 'delta[2]'] != 0, xlab = 'iteration', ylab = 'delta[2] presence',
                    main = 'delta[2] presence')
ts.plot(resultsAFT[ , 'delta[2]'], xlab = 'iterations', ylab = 'delta[2]',
               main = 'hemoglobin coefficient')

## posterior probability of inclusion    
mean(resultsAFT[ , 'delta[2]'] != 0)  
```

# Summary of RJMCMC

- Mixing will generally be better than simply using an indicator function without RJMCMC.
- One can use RJ for variable selection in NIMBLE either with or without indicator functions.
   - Use of indicator functions allows for hierarchical structure for selection of multiple variables.
- Adaptation for coefficient samplers only occurs when the coefficient is "in the model".
- Run time should be much faster *if* posterior probability of inclusion is not large. 
- Tuning parameter of RJ proposal scale (sd) must be chosen.


